Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-10-20 motivation.tex
motivation

As this discussion hints, implementations of SI on real databases
such as PostgresSQL are highly complicated, often running into
thousands of lines of code.  Nonetheless, the semantics of SI, in
terms of how it effects transitions on the database state, can be
captured in a fairly simple model. First, effects induced by one
transaction (call it $T$) are not visible to another concurrently executing one
during $T$'s execution. Thus, from $T$'s perspective,
the global state does not change during its execution.  More formally,
for every operation performed by $T$, the global state $T$ witnesses
before ($\stg$) and after ($\stg')$ executing the operation is the
same ($\stg'=\stg$).  After $T$ finishes execution, it commits its
changes to the actual database, which may have already incorporated
the effects of concurrent transactions. In executions where $T$
successfully commits, concurrent transactions are guaranteed to not be
in write-write conflict with $T$. Thus, if $\stg$ is the global state
that $T$ witnessed when it finished execution (the snapshot state),
and $\stg'$ is the state to which $T$ commits, then the difference
between $\stg$ and $\stg'$ should not result in a write-write conflict
with $T$. To concretize this notion, let the database state be a map
from database locations to values, and let $\stl$ denote a
transaction-local log that maps the locations being written to their
updated values.  The absence of write-write conflicts between $T$ and
the diff between $\stg$ and $\stg'$ can be expressed as: $\forall
x\in\mathit{dom}(\delta)$, $\stg'(x) = \stg(x)$.  In other words, the
semantics of SI can be captured as an axiomatization over transitions
of the database state ($\Delta \longrightarrow \Delta'$) during a
transaction's ($T$) lifetime:
\begin{itemize}
  \item While $T$ executes, $\Delta' = \Delta$.
  \item After $T$ finishes execution, but before it commits its local
    state $\delta$, $\forall(x\in\mathit{dom}(delta)).~\Delta'(x) = \Delta(x)$.
\end{itemize}

@@

We approach the verification problem by first observing that a typical
relational database is a significantly simpler abstraction than shared
memory.  Its primary data structure is a table, with no primitive
support for pointers, linked data structures, or aliasing.  Although a
database essentially abstracts a mutable state, this state is managed
through a well-defined fixed number of interfaces (SQL statements),
each tagged with a logical formula describing what records are
accessed and updated.

@@

Fig.~\ref{fig:foreach_code} shows a (simplified) snippet of code taken
from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
been replaced with constants (\C{k1},\C{k2}, and \C{k3}).  The body of
the loop executes a SQL update followed by an insert.  Recall that a
transaction reads from the global database ($\Delta$), and writes to a
transaction-local database ($\delta$) before committing these
updates. An update statement filters the records that match the search
criteria from $\Delta$ and computes the updated records that are to be
added to the local database. Thus, the state transformer for the
update statement (call it $\T_U$) is the following function on
sets\footnote{$\bind$ has higher precedence than $\cup$.}:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \stg \bind(\lambda s. 
      \itel{{\sf table}(s) = \C{Stock} \conj 
        s.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\\hspace*{1.3in}}
           {\{ \langle s\ {\sf with}\  s_\C{qty} = \C{k1}\rangle \}\\\hspace*{1.3in}}
           {\emptyset})
\end{array}
\end{smathpar}
% \begin{smathpar}
% \begin{array}{c}
%   \lambda\Delta.\lambda\delta.\, \delta \;\cup\; 
%       \Delta \,\bind\, (\lambda s.\, 
%           \C{if}\;{\C{stock}(s) \conj 
%                    \C{s.s\_i\_id} = \C{item\_req.ol\_i\_id}}\\
%            \hspace*{0.9in}
%            \C{then}\;{\{\{s \;\C{with}\; \C{s\_qty} = \C{k1}\}\}}\;
%            \C{else}\;{\emptyset})
% \end{array}
% \end{smathpar}
% \begin{verbatim}
%   Rmem(σ'(s')) = Rmem(σ(s')) U 
%                  Rmem[\x. Rmem[\y. if y<x then {y+x} else {}](σ(s)) U {x}](xs)
% \end{verbatim}
% Observe that
% $\T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
% F_U(\Delta)$.
Here, the set bind operator extracts record elements ($\C{s}$) from
the database, checks the precondition of the update action, and if
satisfied, constructs a new set containing a single record that is
identical to $\C{s}$ except that it binds field $\C{s_{qty}}$ to value
$\C{k_1}$.  This new set is added (via set union) to the existing
local database state $\delta$.\footnote{For now, assume that the
  record being added to the local database has not previously been
  updated.}

The transformer ($\T_I(\delta,\Delta)$) for the subsequent \C{insert}
statement can be similarly constructed:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup
       \{\langle\C{ol\_o\_id}=\C{k2},
                 \C{ol\_d\_id}=\C{k3},\, \C{ol\_i\_id}=\C{item\_req.ol\_i\_id},\, 
                 \C{ol\_qty}=\C{item\_req.ol\_qty}\rangle\}

\end{array}
\end{smathpar}
Observe that both transformers are of the form $T(\delta,\Delta) =
\stl \cup F(\stl,\stg)$, where $F$ is a function that returns the set
of records added  to the transaction-local database
($\stl$). Let $\F_U$ and $\F_I$ be the corresponding functions for
$\T_U$ and $\T_I$ shown above. The state transformation induced by the
loop body in Fig.~\ref{fig:new_order_code} can be expressed as the
following composition of $\F_U$ and $\F_I$:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \F_U(\stl,\stg) \cup \F_I(\stl,\stg)
\end{array}
\end{smathpar}
The transformer for the loop itself can now be computed to be:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \C{item\_reqs}\bind
      (\lambda\C{item\_req}.~ \F_U(\stl,\stg) \cup \F_I(\stl,\stg))
\end{array}
\end{smathpar}
Observe that the structure of the transformer mirrors the structure of
the program itself. In particular, SQL statements become set
operations, and the \C{DB} monad's $\bind$ becoms set monadic $\bind$.
As we demonstrate, the advantage of inferring such transformers is
that we can now make use of a semantics-preserving translation from
the domain of sets equipped with $\bind$ to a decidable fragment of
first-order logic, allowing us to leverage SMT solvers for automated
proofs without having to infer potentially complex thread-local
invariants or intermediate assertions.  Sec.~\ref{sec:inference}
describes this translation. In the exposition thus far, we assumed
$\Delta$ remains invariant, which is clearly not the case when we
admit concurrency.  Necessary concurrency extensions of the state
transformer semantics to deal with interference is also covered in
Sec.~\ref{sec:inference}.  Before presenting the transformer
semantics, we first focus our attention in the following two sections
on the theoretical foundations for weak isolation, upon which this
semantics is based.

>>>>

As this discussion hints, implementations of SI on real databases
such as PostgresSQL are highly complicated, often running into
thousands of lines of code.  Nonetheless, the semantics of SI, in
terms of how it effects transitions on the database state, can be
captured in a fairly simple model. First, effects induced by one
transaction (call it $T$) are not visible to another concurrently executing one
during $T$'s execution. Thus, from $T$'s perspective,
the global state does not change during its execution.  More formally,
for every operation performed by $T$, the global state $T$ witnesses
before ($\stg$) and after ($\stg')$ executing the operation is the
same ($\stg'=\stg$).  After $T$ finishes execution, it commits its
changes to the actual database, which may have already incorporated
the effects of concurrent transactions. In executions where $T$
successfully commits, concurrent transactions are guaranteed to not be
in write-write conflict with $T$. Thus, if $\stg$ is the global state
that $T$ witnessed when it finished execution (the snapshot state),
and $\stg'$ is the state to which $T$ commits, then the difference
between $\stg$ and $\stg'$ should not result in a write-write conflict
with $T$. To concretize this notion, let the database state be a map
from database locations to values, and let $\stl$ denote a
transaction-local log that maps the locations being written to their
updated values.  The absence of write-write conflicts between $T$ and
the diff between $\stg$ and $\stg'$ can be expressed as: $\forall
x\in\mathit{dom}(\delta)$, $\stg'(x) = \stg(x)$.  In other words, the
semantics of SI can be captured as an axiomatization over transitions
of the database state ($\Delta \longrightarrow \Delta'$) during a
transaction's ($T$) lifetime:
\begin{itemize}
  \item While $T$ executes, $\Delta' = \Delta$.
  \item After $T$ finishes execution, but before it commits its local
    state $\delta$, $\forall(x\in\mathit{dom}(\delta)).~\Delta'(x) = \Delta(x)$.
\end{itemize}

@@

We approach the verification problem by first observing that a
relational database is a significantly simpler abstraction than shared
memory. Its primary data structure is a table, with no primitive
support for pointers, linked data structures, or aliasing.  Although a
database essentially abstracts a mutable state, this state is managed
through a well-defined fixed number of interfaces (SQL statements),
each tagged with a logical formula describing what records are
accessed and updated.

@@

Fig.~\ref{fig:foreach_code} shows a (simplified) snippet of code taken
from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
been replaced with constants (\C{k1},\C{k2}, and \C{k3}).  The body of
the loop executes a SQL update followed by an insert.  Recall that a
transaction reads from the global database ($\Delta$), and writes to a
transaction-local database ($\delta$) before committing these
updates. An update statement filters the records that match the search
criteria from $\Delta$ and computes the updated records that are to be
added to the local database. Thus, the state transformer for the
update statement (call it $\T_U$) is the following function on
sets\footnote{$\bind$ has higher precedence than $\cup$; angle braces
($\langle \ldots \rangle$) are used to denote records in math}:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \stg \bind(\lambda s. 
      \itel{{\sf table}(s) = \C{Stock} \conj 
        s.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\\hspace*{1.3in}}
           {\{ \langle \C{s\_i\_id}=s.\C{s\_i\_id};\, 
                       \C{s\_d\_id}=s.\C{s\_d\_id};\,
                       \C{s\_qty} = \C{k1}\rangle \}\\\hspace*{1.3in}}
           {\emptyset})
\end{array}
\end{smathpar}
% \begin{smathpar}
% \begin{array}{c}
%   \lambda\Delta.\lambda\delta.\, \delta \;\cup\; 
%       \Delta \,\bind\, (\lambda s.\, 
%           \C{if}\;{\C{stock}(s) \conj 
%                    \C{s.s\_i\_id} = \C{item\_req.ol\_i\_id}}\\
%            \hspace*{0.9in}
%            \C{then}\;{\{\{s \;\C{with}\; \C{s\_qty} = \C{k1}\}\}}\;
%            \C{else}\;{\emptyset})
% \end{array}
% \end{smathpar}
% \begin{verbatim}
%   Rmem(σ'(s')) = Rmem(σ(s')) U 
%                  Rmem[\x. Rmem[\y. if y<x then {y+x} else {}](σ(s)) U {x}](xs)
% \end{verbatim}
% Observe that
% $\T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
% F_U(\Delta)$.
Here, the set bind operator extracts record elements ($\C{s}$) from
the database, checks the precondition of the update action, and if
satisfied, constructs a new set containing a single record that is
identical to $\C{s}$ except that it binds field $\C{s_{qty}}$ to value
$\C{k_1}$.  This new set is added (via set union) to the existing
local database state $\delta$.\footnote{For now, assume that the
  record being added to the local database has not previously been
  updated.}

The transformer ($\T_I(\delta,\Delta)$) for the subsequent \C{insert}
statement can be similarly constructed:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup
       \{\langle\C{ol\_o\_id}=\C{k2},\,
                 \C{ol\_d\_id}=\C{k3},\, \C{ol\_i\_id}=\C{item\_req.ol\_i\_id},\, 
                 \C{ol\_qty}=\C{item\_req.ol\_qty}\rangle\}

\end{array}
\end{smathpar}
Observe that both transformers are of the form $\T(\delta,\Delta) =
\stl \cup \F(\stl,\stg)$, where $\F$ is a function that returns the
set of records added  to the transaction-local database ($\stl$). Let
$\F_U$ and $\F_I$ be the corresponding functions for $\T_U$ and $\T_I$
shown above. The state transformation induced by the loop body in
Fig.~\ref{fig:new_order_code} can be expressed as the following
composition of $\F_U$ and $\F_I$:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \F_U(\stl,\stg) \cup \F_I(\stl,\stg)
\end{array}
\end{smathpar}
The transformer for the loop itself can now be computed to be:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \C{item\_reqs}\bind
      (\lambda\C{item\_req}.~ \F_U(\stl,\stg) \cup \F_I(\stl,\stg))
\end{array}
\end{smathpar}
Observe that the structure of the transformer mirrors the structure of
the program itself. In particular, SQL statements become set
operations, and the \C{DB} monad's $\bind$ becoms set monadic $\bind$.
As we demonstrate, the advantage of inferring such transformers is
that we can now make use of a semantics-preserving translation from
the domain of sets equipped with $\bind$ to a decidable fragment of
first-order logic, allowing us to leverage SMT solvers for automated
proofs without having to infer potentially complex thread-local
invariants or intermediate assertions.  Sec.~\ref{sec:inference}
describes this translation. In the exposition thus far, we assumed
$\Delta$ remains invariant, which is clearly not the case when we
admit concurrency.  Necessary concurrency extensions of the state
transformer semantics to deal with interference is also covered in
Sec.~\ref{sec:inference}.  Before presenting the transformer
semantics, we first focus our attention in the following two sections
on the theoretical foundations for weak isolation, upon which this
semantics is based.
